---
title: "Younger electorates vote independent in Tunisian Municipal Elections"
author: "Christopher Barrie"
date: 2018-05-23
categories: ["OpenData", "Elections", "Tunisia"]
tags: ["OpenData", "Elections", "Tunisia"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(openxlsx)
library(tidyverse)
library(leaflet)
library(rgdal)
library(ggpubr)
library(corrplot)
library(plotly)
library(gridExtra)

sheets <- openxlsx::getSheetNames("/Users/christopherbarrie/Dropbox/Desktop/R/Tunisia maps/Municipales/raw/tun_muni1.xlsx")

df <- data.frame()

for (x in 1:27) {
  df <- rbind(df, read.xlsx("/Users/christopherbarrie/Dropbox/Desktop/R/Tunisia maps/Municipales/raw/tun_muni1.xlsx", sheet=sheets[x]))
}

names(df) <- c("name.fr","name_ar", 
               "type_list", "name_list", 
               "name_list_ar", "votes", 
               "votes_prop", "seats", 
               "time_stamp", "source", 
               "delegation")

# get election data

df_EL_short <- df %>% 
  select(votes, votes_prop, seats, type_list, name_list, name.fr, delegation)

# generate vote percentage
df_EL_short$votes_percent <- df_EL_short$votes_prop*100
# change to two decimal places
df_EL_short$votes_percent <- formatC(df_EL_short$votes_percent, digits = 2, format = "f")
# convert vote percent to numeric
df_EL_short$votes_percent <- as.numeric(df_EL_short$votes_percent)
# convert name to character
df_EL_short$name.fr <- as.character(df_EL_short$name.fr)

# get votes by name.fr
df_Votes <-  df_EL_short %>%
  group_by(name.fr) %>%
  summarize(sum_votes = sum(votes))

df_Ages <- read.xlsx("/Users/christopherbarrie/Dropbox/Desktop/R/Tunisia maps/Municipales/muni_ages.xlsx", 
                     sheet="Sheet1")

df_Votes_Ages <- merge(df_Votes, df_Ages, by="name.fr")

# check turnout
#(sum(df_totals$sum_votes, na.rm=TRUE)/sum(df_totals$total, na.rm=TRUE))*100
# add turnout to data.frame
df_Votes_Ages$turnout <- (df_Votes_Ages$sum_votes/df_Votes_Ages$total)*100

# generate population ages as percentage of total

df_Votes_Ages$pop1835 <- (df_Votes_Ages$`18_to_35`/df_Votes_Ages$total)*100
df_Votes_Ages$pop3640 <- (df_Votes_Ages$`36_to_40`/df_Votes_Ages$total)*100
df_Votes_Ages$pop4160 <- (df_Votes_Ages$`41_to_60`/df_Votes_Ages$total)*100
df_Votes_Ages$pop60p <- (df_Votes_Ages$`60_plus`/df_Votes_Ages$total)*100
```

# TL;DR

Municipalities with a greater number of younger registered voters saw a higher vote share for independent lists. Similarly, younger electorates were less likely to vote for established parties.

## Introduction

On May 6, Tunisia held its first municipal elections since the 2011 Revolution. In a [previous post](https://cjbarrie.netlify.com/post/bpost18062018/), I showed that younger municipalities and municipalities with more political competition/electoral mobilization saw higher turnout. In this post, I look at associations between municipality-level demographics and voting patterns again but focus on the the types of parties that were successful in each municipality. Once again, I make use of the data made available through the [online community-led efforts](https://www.huffpostmaghreb.com/entry/le-crowdsourcing-en-tunisie-pourrait-il-devenir-une-reference-en-matiere-de-donnees-publiques-cahier-de-la-liberte-y-repond_mg_5af443e6e4b0859d11d0ef8a?utm_hp_ref=mg-elections-municipales-tunisie) of a team of Tunisian OpenData organizations. I combine these data with municipality-level demographic data [provided by the ISIE](http://www.isie.tn/elections/elections-municipales-2018/statistiques/) in order to investigate voting patterns.

## Age and votes for independents

The first local elections to be held in post-revolution Tunisia were particularly notable for the success of independent lists. In all, independent lists managed to win some 32% of the overall vote, beating into second and third place established political forces such as Ennahdha and Nidaa Tounes. There was an astounding proliferation of such lists, the large majority of which only contested one seat. The table below details each of the more than 600 individual independent lists, as well as the number of seats they contested. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
library(kableExtra)
df_EL_short$obs <- rep(1, nrow(df_EL_short))
df_IN <- df_EL_short %>%
  filter(type_list == "Liste indépendante") %>%
  group_by(name_list) %>%
  summarize(Number_seats_contested = sum(obs))
colnames(df_IN)[1] <- paste0("Independent list name")
colnames(df_IN)[2] <- paste0("Number of seats contested")

kable(df_IN) %>%
kable_styling() %>%
scroll_box(height = "500px")
```
\  

Below, I look at bivariate associations between municipality age demographics and the observed vote share for independents lists. The y-axis thus measures the sum of the shares of the vote achieved by independent lists in each municipality.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# get vote percentages for independents
df_Elec_IN <- df_EL_short %>%
  filter(type_list == "Liste indépendante") %>%
  select(votes_percent, name.fr) %>%
  group_by(name.fr) %>%
  summarize(IN_votes_percent = sum(votes_percent))

# merge
df_Votes_Ages1 <- merge(df_Votes_Ages, df_Elec_IN, by="name.fr", all=TRUE)

agein1 <- ggplot(df_Votes_Ages1, aes(x=pop1835, y=IN_votes_percent)) +
  geom_point(alpha=0.5, color="#c0392b") +
  geom_hline(yintercept=1, size=0.4, color="black") +
  geom_smooth(method= lm, alpha=0.25, color="#c0392b", fill="#c0392b") +
  theme_minimal() + theme(aspect.ratio=1) +
  labs(x="18-35 %", y="Independent %", title="")

agein2 <- ggplot(df_Votes_Ages1, aes(x=pop60p, y=IN_votes_percent)) +
  geom_point(alpha=0.5, color="#c0392b") +
  geom_hline(yintercept=1, size=0.4, color="black") +
  geom_smooth(method= lm, alpha=0.25, color="#c0392b", fill="#c0392b") +
  theme_minimal() + theme(aspect.ratio=1) +
  labs(x="60+  %", y="Independent %", title="")

grid.arrange(agein1, agein2, ncol=2)
```

We see here that independents seem to gain a greater share of the vote in municipalities with younger electorates. What does this picture look like when we look not at independent parties but at the vote share of established parties? Below I show associations of vote share for established parties with the same age demographic measures analysed above. Established parties (and coalitions of parties e.g., Front Populaire) include all lists that gained at least 1% of the vote share nationally and are not independent lists. These included: Ennahdha; Nidaa Tounes; Machrouu Tounes; Courant Démocrate; Parti Destourien Libre Pdl; Front Populaire; Union Civile; Al Irada (Mouvement Tunisie Volonté); Mouvement Du Peuple; and Afek Tounes. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# get vote percentages for Established parties
df_Elec_EST <- df_EL_short %>%
  filter(name_list == "Ennahdha" | name_list == "Nidaa Tounes"
         | name_list == "Machrouu Tounes" | name_list == "Courant Démocrate"
         | name_list == "Parti Destourien Libre Pdl" | name_list == "Front Populaire"
         | name_list == "Union Civile" | name_list == "AL IRADA"
         | name_list == "Mouvement Du Peuple" | name_list == "Afek Tounes") %>%
  select(votes_percent, name.fr) %>%
  group_by(name.fr) %>%
  summarize(EST_votes_percent = sum(votes_percent))

# merge
df_Votes_Ages2 <- merge(df_Votes_Ages, df_Elec_EST, by="name.fr", all=TRUE)

ageest1 <- ggplot(df_Votes_Ages2, aes(x=pop1835, y=EST_votes_percent)) +
  geom_point(alpha=0.5, color="#c0392b") +
  geom_hline(yintercept=1, size=0.4, color="black") +
  geom_smooth(method= lm, alpha=0.25, color="#c0392b", fill="#c0392b") +
  theme_minimal() + theme(aspect.ratio=1) +
  labs(x="18-35 %", y="Established parties %", title="")

ageest2 <- ggplot(df_Votes_Ages2, aes(x=pop60p, y=EST_votes_percent)) +
  geom_point(alpha=0.5, color="#c0392b") +
  geom_hline(yintercept=1, size=0.4, color="black") +
  geom_smooth(method= lm, alpha=0.25, color="#c0392b", fill="#c0392b") +
  theme_minimal() + theme(aspect.ratio=1) +
  labs(x="60+  %", y="Established parties %", title="")

grid.arrange(ageest1, ageest2, ncol=2)
```

A similar pattern emerges here: it appears that younger electorates are less likely to vote for established political forces, while more aging electorates are more likely to support established political parties. 

Overall, then, if the Tunisian Municipal Elections were notable for the success of independents, it seems that young voters played a part in their success. What is more, younger electorates seem to be less likely to support established political forces.